---
title: "Multiple Metric Learning Methods in scLearn"
package: "scLearn"
author:
  - name: Bin Duan
    affiliation: SJTU
    email: binduan@sjtu.edu.cn
date: "`r Sys.Date()`"
output: 
  BiocStyle::html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    fig_width: 8
    fig_height: 6
  pdf_document:
    toc: true
vignette: >
  %\VignetteIndexEntry{Metric Learning in scLearn}
  %\VignetteEngine{knitr::rmarkdown}
  %\usepackage[utf8]{inputenc}  
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center",
  cache = TRUE
)
library(BiocStyle)
```


## Introduction

Comprehensive summary table of the six metric learning methods used in scLearn

| **Method**                                            | **Core Idea**                                                | **Advantages**                                               | **Disadvantages**                                            | **Recommended Sample Size** |
| ----------------------------------------------------- | ------------------------------------------------------------ | ------------------------------------------------------------ | ------------------------------------------------------------ | --------------------------- |
| **DCA** (Discriminative Component Analysis)           | Linear transformation maximizing the between-class / within-class variance ratio (Fisher-style) | Fast, simple, stable; no iterative optimization              | Assumes linear class separability; not suitable for complex manifolds | Small to medium             |
| **LMNN** (Large Margin Nearest Neighbors)             | Learns a transformation that maintains class-consistent k-NN relationships with margin constraints | High classification accuracy; interpretable                  | Computationally expensive; requires many triplets; slow without parallelization | Small to medium             |
| **MSL-LMNN** (Multi-Similarity Loss-Based Metric Learning) | Enhances LMNN by leveraging multiple similarity cues and continuous loss instead of discrete triplets | More stable optimization; better generalization with fewer pairs | Still requires pairwise comparisons; moderately slow without batching | Medium to large             |
| **NCA** (Neighborhood Components Analysis)            | Maximizes the probability of correct classification under stochastic nearest neighbors | Theoretically elegant; good embedding quality                | Sensitive to noise; no margin; costly without approximate neighbors | Small to medium             |
| **ITML** (Information-Theoretic Metric Learning)      | Learns Mahalanobis distance with relative constraints, regularized by information-theoretic divergence | Robust to noise; stable convergence; works with few constraints | Requires careful tuning; slow on high-dimensional data       | Small to medium             |
| **MMC** (Maximum Margin Clustering)                   | Maximizes inter-class separation while minimizing intra-class spread using scatter matrices | Good for clustering and large-margin separation              | High memory usage; expensive eigen decomposition; difficult on large datasets | Medium (with memory)        |



## Installation

```{r}
library(scLearn)
library(tidyverse) 
library(SingleCellExperiment)
library(M3Drop)
```


## Data preparation

+ `Reference Cell Database`: baron-human

+ `Query Cell Data`: Muraro-human

```{r}
data(RefCellData)
data(QueryCellData)
stratified_sample <- function(cell_types, n_total = 300, min_per_type = 10) {
  sampled <- lapply(split(seq_along(cell_types), cell_types), function(idx) {
    if(length(idx) > min_per_type) {
      sample(idx, max(min_per_type, round(n_total * length(idx)/length(cell_types))))
    } else {
      idx
    }
  })
   unlist(sampled)
}


set.seed(123)

cell_types <- colData(RefCellData)$cell_type1
sampled_cells <- stratified_sample(cell_types)
RefCellData <- RefCellData[, sampled_cells]
RefCellData

cell_types <- colData(QueryCellData)$cell_type1
sampled_cells <- stratified_sample(cell_types)
QueryCellData <- QueryCellData[, sampled_cells]
QueryCellData
```

## Quality control

```{r}
RefRawcounts <- assays(RefCellData)[[1]]
RefAnn <- as.character(RefCellData$cell_type1)
names(RefAnn) <- colnames(RefCellData)

RefDataQC <- Cell_qc(
  expression_profile = RefRawcounts, 
  sample_information_cellType = RefAnn, 
  sample_information_timePoint = NULL, 
  species = "Hs",
  gene_low = 500, 
  gene_high = 10000, 
  mito_high = 0.1, 
  umi_low = 1500, 
  umi_high = Inf, 
  logNormalize = TRUE, 
  plot = FALSE, 
  plot_path = "./quality_control.pdf")
```

## Filtering rare cell

```{r}
RefDataQC_filtered <- Cell_type_filter(
  expression_profile = RefDataQC$expression_profile, 
  sample_information_cellType = RefDataQC$sample_information_cellType,
  sample_information_timePoint = NULL,
  min_cell_number = 10)
```

## Feature selection

highest variable genes

```{r}
RefDataQC_HVG_names <- Feature_selection_M3Drop(
  expression_profile = RefDataQC_filtered$expression_profile,
  log_normalized = TRUE, 
  threshold = 0.05)
```

## QueryData Processing

```{r}
QueryRawcounts <- assays(QueryCellData)[[1]]
QueryDataQC <- Cell_qc(
  expression_profile = QueryRawcounts, 
  species = "Hs", 
  gene_low = 50, 
  umi_low = 50)

rownames(QueryDataQC$expression_profile) <- gsub("__\\w+\\d+", "", rownames(QueryDataQC$expression_profile))

QueryData_trueLabel <- as.character(QueryCellData$cell_type1)
names(QueryData_trueLabel) <- colnames(QueryCellData)
```


## DCA

Discriminative Component Analysis (DCA) Transformation

+ model learning

```{r}
scLearn_model_learning_result_DCA <- scLearn_model_learning(
  high_varGene_names = RefDataQC_HVG_names,
  expression_profile = RefDataQC_filtered$expression_profile,
  sample_information_cellType = RefDataQC_filtered$sample_information_cellType,
  sample_information_timePoint = NULL,
  method = "DCA",
  bootstrap_times = 1,
  cutoff = 0.01,
  dim_para = 0.999
)
```

**Results**：`scLearn_model_learning_result_DCA`is the final result file containing the processed reference cell matrix.

-   `high_varGene_names`: The set of highly variable genes (`r length(scLearn_model_learning_result_DCA$high_varGene_names)` HVGs)

-   `simi_threshold_learned`: The similarity results between cell types

-   `trans_matrix_learned`: The transformed matrix (`r dim(scLearn_model_learning_result_DCA$trans_matrix_learned[[1]])[1]` features \* `r dim(scLearn_model_learning_result_DCA$trans_matrix_learned[[1]])[2]` HVGs)

-   `feature_matrix_learned`: After DCA, the matrix of cell type features (these features are obtained by dimensionality reduction of the highly variable gene set) (`r dim(scLearn_model_learning_result_DCA$feature_matrix_learned[[1]])[1]` celltype \* `r dim(scLearn_model_learning_result_DCA$feature_matrix_learned[[1]])[2]` features)


+ cell assignment

```{r}
scLearn_predict_result_DCA <- scLearn_cell_assignment(
  scLearn_model_learning_result = scLearn_model_learning_result_DCA, 
  expression_profile_query = QueryDataQC$expression_profile, 
  diff = 0.05, 
  threshold_use = TRUE, 
  vote_rate = 0.6)

head(scLearn_predict_result_DCA)
```

**Results**: The output consists of three columns of information. The `Query_cell_id` represents the cell ID, `Predict_cell_type` indicates the predicted cell type, and `Additional_information` provides the similarity results. Cells with a `Predict_cell_type` of *unassigned* are those that do not intersect with the reference cell set.

+ Accuracy

```{r}
QueryData_CellType_DCA <- QueryData_trueLabel |> 
  as.data.frame() |>
  tibble::rownames_to_column("CellID") |>
  dplyr::rename(trueLabel = QueryData_trueLabel) |>  
  dplyr::inner_join(scLearn_predict_result_DCA |>
                      as.data.frame() |>
                      dplyr::rename(predLabel = Predict_cell_type),
                    by = c("CellID" = "Query_cell_id")) |>
  dplyr::select(CellID, trueLabel, predLabel, everything())

print(
  paste("DCA Accuracy =", 
    sprintf("%1.2f%%",
      100 * sum(QueryData_CellType_DCA$predLabel == QueryData_CellType_DCA$trueLabel) / nrow(QueryData_CellType_DCA))))
```


## LMNN

Implements Large Margin Nearest Neighbor (LMNN) metric learning Transformation

+ model learning

```{r}
scLearn_model_learning_result_LMNN <- scLearn_model_learning(
  high_varGene_names = RefDataQC_HVG_names,
  expression_profile = RefDataQC_filtered$expression_profile,
  sample_information_cellType = RefDataQC_filtered$sample_information_cellType,
  sample_information_timePoint = NULL,
  method = "LMNN",
  bootstrap_times = 1,
  cutoff = 0.01,
  dim_para = 0.999,
  k = 5,
  max_iter = 1
)
```


+ cell assignment

```{r}
scLearn_predict_result_LMNN <- scLearn_cell_assignment(
  scLearn_model_learning_result = scLearn_model_learning_result_LMNN, 
  expression_profile_query = QueryDataQC$expression_profile, 
  diff = 0.05, 
  threshold_use = TRUE, 
  vote_rate = 0.6)

head(scLearn_predict_result_LMNN)
```

+ Accuracy

```{r}
QueryData_CellType_LMNN <- QueryData_trueLabel |> 
  as.data.frame() |>
  tibble::rownames_to_column("CellID") |>
  dplyr::rename(trueLabel = QueryData_trueLabel) |>  
  dplyr::inner_join(scLearn_predict_result_LMNN |>
                      as.data.frame() |>
                      dplyr::rename(predLabel = Predict_cell_type),
                    by = c("CellID" = "Query_cell_id")) |>
  dplyr::select(CellID, trueLabel, predLabel, everything())

print(
  paste("LMNN Accuracy =", 
    sprintf("%1.2f%%",
      100 * sum(QueryData_CellType_LMNN$predLabel == QueryData_CellType_LMNN$trueLabel) / nrow(QueryData_CellType_LMNN))))
```


## MSL-LMNN

Implements Multi-Similarity Loss-Based Large Margin Nearest Neighbor (LMNN) metric learning Transformation

**Note: MSL-LMNN took a long time**

+ model learning

```{r}
scLearn_model_learning_result_MSL <- scLearn_model_learning(
  high_varGene_names = RefDataQC_HVG_names,
  expression_profile = RefDataQC_filtered$expression_profile,
  sample_information_cellType = RefDataQC_filtered$sample_information_cellType,
  sample_information_timePoint = NULL,
  method = "MSL",
  bootstrap_times = 1,
  cutoff = 0.01,
  dim_para = 0.999,
  max_iter = 1
)
```


+ cell assignment

```{r}
scLearn_predict_result_MSL <- scLearn_cell_assignment(
  scLearn_model_learning_result = scLearn_model_learning_result_MSL, 
  expression_profile_query = QueryDataQC$expression_profile, 
  diff = 0.05, 
  threshold_use = TRUE, 
  vote_rate = 0.6)

head(scLearn_predict_result_MSL)
```

+ Accuracy

```{r}
QueryData_CellType_MSL <- QueryData_trueLabel |> 
  as.data.frame() |>
  tibble::rownames_to_column("CellID") |>
  dplyr::rename(trueLabel = QueryData_trueLabel) |>  
  dplyr::inner_join(scLearn_predict_result_MSL |>
                      as.data.frame() |>
                      dplyr::rename(predLabel = Predict_cell_type),
                    by = c("CellID" = "Query_cell_id")) |>
  dplyr::select(CellID, trueLabel, predLabel, everything())

print(
  paste("MSL Accuracy =", 
    sprintf("%1.2f%%",
      100 * sum(QueryData_CellType_MSL$predLabel == QueryData_CellType_MSL$trueLabel) / nrow(QueryData_CellType_MSL))))
```

## NCA

Neighborhood Components Analysis (maximizes leave-one-out classification probability)

+ model learning

```{r}
scLearn_model_learning_result_NCA <- scLearn_model_learning(
  high_varGene_names = RefDataQC_HVG_names,
  expression_profile = RefDataQC_filtered$expression_profile,
  sample_information_cellType = RefDataQC_filtered$sample_information_cellType,
  sample_information_timePoint = NULL,
  method = "NCA",
  bootstrap_times = 1,
  cutoff = 0.01,
  dim_para = 0.999,
  max_iter = 1
)
```

+ cell assignment

```{r}
scLearn_predict_result_NCA <- scLearn_cell_assignment(
  scLearn_model_learning_result = scLearn_model_learning_result_NCA, 
  expression_profile_query = QueryDataQC$expression_profile, 
  diff = 0.05, 
  threshold_use = TRUE, 
  vote_rate = 0.6)

head(scLearn_predict_result_NCA)
```

+ Accuracy

```{r}
QueryData_CellType_NCA <- QueryData_trueLabel |> 
  as.data.frame() |>
  tibble::rownames_to_column("CellID") |>
  dplyr::rename(trueLabel = QueryData_trueLabel) |>  
  dplyr::inner_join(scLearn_predict_result_NCA |>
                      as.data.frame() |>
                      dplyr::rename(predLabel = Predict_cell_type),
                    by = c("CellID" = "Query_cell_id")) |>
  dplyr::select(CellID, trueLabel, predLabel, everything())

print(
  paste("NCA Accuracy =", 
    sprintf("%1.2f%%",
      100 * sum(QueryData_CellType_NCA$predLabel == QueryData_CellType_NCA$trueLabel) / nrow(QueryData_CellType_NCA))))
```

## ITML

Information-Theoretic Metric Learning

+ model learning

```{r}
scLearn_model_learning_result_ITML <- scLearn_model_learning(
  high_varGene_names = RefDataQC_HVG_names,
  expression_profile = RefDataQC_filtered$expression_profile,
  sample_information_cellType = RefDataQC_filtered$sample_information_cellType,
  sample_information_timePoint = NULL,
  method = "ITML",
  bootstrap_times = 1,
  cutoff = 0.01,
  dim_para = 0.999,
  max_iter = 1
)
```

+ cell assignment

```{r}
scLearn_predict_result_ITML <- scLearn_cell_assignment(
  scLearn_model_learning_result = scLearn_model_learning_result_ITML, 
  expression_profile_query = QueryDataQC$expression_profile, 
  diff = 0.05, 
  threshold_use = TRUE, 
  vote_rate = 0.6)

head(scLearn_predict_result_ITML)
```

+ Accuracy

```{r}
QueryData_CellType_ITML <- QueryData_trueLabel |> 
  as.data.frame() |>
  tibble::rownames_to_column("CellID") |>
  dplyr::rename(trueLabel = QueryData_trueLabel) |>  
  dplyr::inner_join(scLearn_predict_result_ITML |>
                      as.data.frame() |>
                      dplyr::rename(predLabel = Predict_cell_type),
                    by = c("CellID" = "Query_cell_id")) |>
  dplyr::select(CellID, trueLabel, predLabel, everything())

print(
  paste("ITML Accuracy =", 
    sprintf("%1.2f%%",
      100 * sum(QueryData_CellType_ITML$predLabel == QueryData_CellType_ITML$trueLabel) / nrow(QueryData_CellType_ITML))))
```

## MMC

Maximum Margin Clustering (explicitly optimizes within-class/between-class scatter matrices)

+ model learning

```{r}
scLearn_model_learning_result_MMC <- scLearn_model_learning(
  high_varGene_names = RefDataQC_HVG_names,
  expression_profile = RefDataQC_filtered$expression_profile,
  sample_information_cellType = RefDataQC_filtered$sample_information_cellType,
  sample_information_timePoint = NULL,
  method = "MMC",
  bootstrap_times = 1,
  cutoff = 0.01,
  dim_para = 0.999,
  max_iter = 1
)
```

+ cell assignment

```{r}
scLearn_predict_result_MMC <- scLearn_cell_assignment(
  scLearn_model_learning_result = scLearn_model_learning_result_MMC, 
  expression_profile_query = QueryDataQC$expression_profile, 
  diff = 0.05, 
  threshold_use = TRUE, 
  vote_rate = 0.6)

head(scLearn_predict_result_MMC)
```

+ Accuracy

```{r}
QueryData_CellType_MMC <- QueryData_trueLabel |> 
  as.data.frame() |>
  tibble::rownames_to_column("CellID") |>
  dplyr::rename(trueLabel = QueryData_trueLabel) |>  
  dplyr::inner_join(scLearn_predict_result_MMC |>
                      as.data.frame() |>
                      dplyr::rename(predLabel = Predict_cell_type),
                    by = c("CellID" = "Query_cell_id")) |>
  dplyr::select(CellID, trueLabel, predLabel, everything())

print(
  paste("MMC Accuracy =", 
    sprintf("%1.2f%%",
      100 * sum(QueryData_CellType_MMC$predLabel == QueryData_CellType_MMC$trueLabel) / nrow(QueryData_CellType_MMC))))
```


## Summary

| **Method**                                            | Accuracy(%) |
| ----------------------------------------------------- | ----------- |
| **DCA** (Discriminative Component Analysis)           | 81.23       |
| **LMNN** (Large Margin Nearest Neighbors)             | 79.94       |
| **MSL-LMNN** (Multi-Similarity Loss-Based Metric Learning) | 78.32       |
| **NCA** (Neighborhood Components Analysis)            | 79.94       |
| **ITML** (Information-Theoretic Metric Learning)      | 79.94       |
| **MMC** (Maximum Margin Clustering)                   | 71.20       |


The table compares the accuracy of six metric learning methods, with DCA (Discriminative Component Analysis) achieving the highest accuracy at 81.23%. LMNN, NCA, and ITML share identical performance at 79.94%, followed by MSL at 78.32%. MMC (Maximum Margin Clustering) shows the lowest accuracy among the evaluated methods at 71.20%. (**RefCellData** & **QueryCellData**)

Key observations:

- DCA outperforms all other methods by a clear margin (1.29% higher than the second-best group).
- Three methods (LMNN, NCA, ITML) demonstrate identical performance levels.
- MMC trails significantly behind other approaches, with a ~8-10% accuracy gap compared to top performers.


## System information

```{r}
sessionInfo()
```

